### 2022-02-17 chapter2/chapter.tex
first draft of the whole thing! :D

%!TEX root = ../thesis.tex
\chapter{Approximating fluid queues with the discontinuous Galerkin method \label{ch:galerkin}}

In this chapter we introduce the discontinuous Galerkin (DG) method as applied to fluid queues and to approximate the operator-analytic expressions for fluid-fluid queues in \cite{bo2014}. 
\begin{center}
	\begin{minipage}{0.8\textwidth}
		\textit{Apart from Section~\ref{sec: slope limiting}, this chapter has been taken from Section~4 and Appendix~1 of \cite{blnos2022} with only minor changes, such as notations, so that this chapter is consistent with the rest of the thesis. I am a co-author of the paper \cite{blnos2022}. %The conceptualisation of \cite{blnos2022} was originally by Vikram Sunkara, Nigel Bean and Giang Nguyen, and the original coding was done by Vikram Sunkara. I made significant contributions to Section~3 of the paper, expressing the operator-theoretic expressions to use the same partition as the approximation scheme. I contributed Sections~4.4 and 5.1. I extended the numerical experiments in Section~6 to higher orders and made all the plots in Section~6. Appendix~A is also my original work. I did a significant proportion of the writing of the manuscript and addressed the reviewers comments and also developed code for the numerical experiments.
		}
	\end{minipage}
	\end{center}
% The DG method for B 
    % direct copy paste from the paper
% Problems with DG, non-neg/oscillations
    % solutions: filtering and limiting
% Approximating R: projection, interpolation, cell averages (eluding to QBD-RAP and unif method)
% Approximating Psi (and the stationary distribution, but this isnt really necessary)
    % constructing D
    % solving the matrix-Riccati equation, iteration/algorithm.

@@

% \section{Discontinuous Galerkin Approximation of the Generator of a Fluid Queue}
	% \label{sec:DG}
Discontinuous Galerkin (DG) methods can be used to approximate the solutions to systems of partial differential equations (PDEs). For a more thorough description of these methods see \citep{nodalDGBook}. The domain of approximation is partitioned into intervals, referred to individually as \textit{cells} and collectively as a \textit{mesh}. On each cell, we have a finite element approximation, which constructs a finite-dimensional smooth Sobolev space using piecewise-polynomial basis functions, and then projects the partial differential equations onto this space. This projection leads to a new system of equations, referred to as the \textit{weak form} of the original system of PDEs. Next, we must approximate the \textit{flux operator} which moves probability from one cell to another, in a manner similar to the underlying principle of a finite-volume approximation. This method conserves probability, and can handle discontinuities, such as jumps and point masses. Here we construct the DG approximation to the matrix of operators \(\mathbb B\) which we use later to construct a DG approximation to \(\mathbb D(s)\) then \(\mathbb\Psi(s)\), and ultimately the stationary distribution of an SFFM.

\section{The Partial Differential Equation}
We start by introducing the PDE from which we will extract the approximation to the generator \(\mathbb B\).

Let $f_i(x,t)$ be the joint density of $\{(X_t, \varphi_t)\}$: 
	% 
	\begin{align*} 
	    f_i(x,t) := \cfrac{\partial}{\partial x} 	\mathbb{P}\left(X_t \leq x, \varphi_t = i\right),\quad 0<x<b,\, i \in\calS,
	\end{align*} 
which satisfies the system of partial differential equations 
%
\begin{align*}
\cfrac{\partial}{\partial t} f_i(x,t)& = \sum_{j\in \mathcal{S}}  f_j(x,t)T_{ji} - c_i \cfrac{\partial}{\partial x} f_i(x,t), \quad 0<x<b,\, i\in\mathcal S,
\end{align*}
% 
subject to suitable boundary conditions. In matrix form, 
\begin{align}
\cfrac{\partial}{\partial t} \boldsymbol f(x,t) &= \boldsymbol f(x,t)\bs T -  \cfrac{\partial}{\partial x}\boldsymbol f(x,t)\bs C\label{eqn:pde_density_matrix}, 
\end{align}
where \(\boldsymbol f(x,t) = \left(f_i(x,t)\right)_{i\in\mathcal S}\) is a row-vector. 
% This system of PDEs is closely related to the generator \(\mathbb B\); \(\boldsymbol \mu(t)\) satisfies the operator differential equation
% \begin{align*}
% 	\cfrac{\wrt}{\wrt t}\boldsymbol \mu(t)(\wrt x)  = \boldsymbol \mu(t)\mathbb B(\wrt x) = \boldsymbol f(x,t)\bs T\wrt x - \cfrac{\partial}{\partial x}\boldsymbol f(x,t)\bs C\wrt x,
% \end{align*}
% on the interior of the space \([0,\infty)\). Thus, by approximating the operator on the right-hand side of Equation (\ref{eqn:pde_density_matrix}) we can approximate the infinitesimal operator \(\mathbb B\). The DG method does exactly this, by approximating the operator with a matrix.

\section{Cells, Test Functions, and Weak Formulation}
To begin with, consider an unbounded first fluid level \(\{\ddot X_t,t\geq0\}\), \(\ddot X_t\in(-\infty,\infty)\). We will eventually truncate this space so that we have a finite dimensional approximation; however, this requires a discussion on boundary conditions which we save for later. Let \(\mathcal D_k = [y_k,y_{k+1}],\, k\in\mathbb Z\) partition the domain \((-\infty,\infty)\). We call the \(\calD_k\) \textit{cells}.

On each cell \(\calD_k\) we choose \(N_k\) linearly independent functions \(\{\phi^r_k\}_{r=1}^{N_k}\), compactly supported on \(\calD_k\) (i.e.~\(\phi^r_k(x)=0\) for \(x\notin\calD_k\)) to form a basis for the space \(W_k\), in which we formulate the approximation. Here, as is standard in DG methods \citep{nodalDGBook}, we take \(\{\phi^r_k\}_{r=1}^{N_k}\) to be the space of polynomials of degree \(N_k-1\). It is convenient in this work to take \(\{\phi^r_k\}_{r=1}^{N_k}\) as a basis of Lagrange interpolating polynomials defined by the Gauss-Lobatto quadrature points, since our approximations inherit nice properties from this \citep{nodalDGBook}. However, the constructions presented here are general, and any basis can be used. For the sake of illustration, the reader may think of \(\{\phi^r_k\}_{r=1}^{N_k}\) as the Lagrange polynomials. On each cell \(\mathcal D_k\) we approximate 
\[f_i(x,t)\approx u_{k,i}(x,t)=\sum\limits_{r=1}^{N_k}a_{k,i}^r(t)\phi^r_k(x),\] 
where \(a_{k,i}^r(t)\) are yet-to-be-determined time-dependent coefficients. We refer to \(u_{k,i}\) as the \textit{local} approximation on \(\calD_k\), while the \textit{global} approximation is given by \(\sum\limits_{k\in\mathbb Z}u_{k,i}\) on the whole domain. The whole approximation space is \(\bigoplus\limits_{k\in\mathbb Z} W_k\).

@@

The DG method proceeds by first considering the \textit{weak-formulation} of the PDE, which is constructed from the \textit{strong-form} of the PDE, Equation (\ref{eqn:pde_density_matrix}). In general, to construct the weak-form we need a set of test functions, say \(W\). Now, take the strong form of the PDE, multiply it by some test function \(\psi(x)\in W\), integrate with respect to \(x\), and apply integration by parts to the derivative with respect to \(x\), to get 
\begin{align}
\begin{multlined}[t]
	\int_{x\in\mathbb R}\cfrac{\partial}{\partial t} f_j(x,t)\psi(x)\wrt x = \int_{x\in\mathbb R} \sum_{i\in\calS}f_i(x,t)T_{ij}\psi(x)\wrt x 
	%
	+  \int_{x\in\mathbb R} f_j(x,t)c_j\cfrac{\wrt}{\wrt x}\psi(x)\wrt x \\{}- [ f_j(x,t)c_j\psi(x)]_{x=-\infty}^{x=\infty}, \end{multlined}\label{eqn:weak form}
\end{align}
for \(j\in\calS\). It is common to choose \(W\) such that \(\psi(-\infty)=\psi(\infty)=0\), in which case the last term on the right is zero. Requiring (\ref{eqn:weak form}) to hold for every \(\psi\in W\) gives the weak-formulation of the PDE. For a sufficiently rich set of test functions \(W\) the weak and strong forms of the PDE are equivalent. Solutions to (\ref{eqn:weak form}) are known as \textit{weak} solutions and generalise the concept of a solution of the PDE. For example, this may allow discontinuities with respect to \(x\) in the solution -- something which is ill-defined for the strong form.

For the purpose of DG, we take the set of test functions to be \(W = \bigoplus\limits_{k\in\mathbb Z} W^k\), the same as the set of basis functions of our solution space. Proceeding as described above, the weak formulation is 
\begin{align*}
	\begin{multlined}[t]\int_{x\in\calD_k}\cfrac{\partial}{\partial t} f_j(x,t)\phi^r_k(x)\wrt x = \int_{x\in\calD_k}\sum_{i\in\calS} f_i(x,t)T_{ij}\phi^r_k(x)\wrt x  
	%
	+  \int_{x\in\calD_k} f_j(x,t)c_j\cfrac{\wrt}{\wrt x}\phi^r_k(x)\wrt x \\ {}- [f_j(x,t)c_j\phi^r_k(x)]_{x=y_k}^{x=y_{k+1}}, \end{multlined}
\end{align*}
since \(\phi^r_k\) is compactly supported on \(\calD_k\), for all \(j\in\mathcal S,\,r\in\mathcal N_k\), \(k\in\mathbb Z.\) Now, note that any function \(g(x)\) can be decomposed as \(g(x) = g^{W}(x)+g^\perp(x)\) where \(g^{W}\in W\) and \(g^\perp \in W^\perp\), and \(W^\perp\) is the orthogonal complement of \(W\). Since \(g^\perp\) is orthogonal to \(W\), \(\displaystyle\int_{x}g^\perp(x)\phi^r_k(x)\wrt x=0\) for \(r\in\mathcal N_k,\,k\in\mathbb Z\). Also, note that \(\cfrac{\wrt}{\wrt x}\phi^r_k(x)\in W\). Using this, we can write 
\begin{align*}
	\begin{multlined}[t]\int_{x\in\calD_k}\cfrac{\partial}{\partial t} \left(f_j^W(x,t)+f_j^\perp(x,t)\right)\phi^r_k(x)\wrt x 
	= \int_{x\in\calD_k}\sum_{i\in\calS} \left(f_i^W(x,t)+f_i^\perp(x,t)\right)T_{ij}\phi^r_k(x)\wrt x  
	%
	\\ {}
	+  \int_{x\in\calD_k} \left(f_j^W(x,t)+f_j^\perp(x,t)\right)c_j\cfrac{\wrt}{\wrt x}\phi^r_k(x)\wrt x - [f_j(x,t)c_j\phi^r_k(x)]_{x=y_k}^{x=y_{k+1}}, \end{multlined}
\end{align*}
	which is equivalent to
\begin{align}
	\begin{multlined}[t]\int_{x\in\calD_k}\cfrac{\partial}{\partial t} f_j^W(x,t)\phi^r_k(x)\wrt x = \int_{x\in\calD_k}\sum_{i\in\calS} f_i^W(x,t)T_{ij}\phi^r_k(x)\wrt x  
	%
	+  \int_{x\in\calD_k} f_j^W(x,t)c_j\cfrac{\wrt}{\wrt x}\phi^r_k(x)\wrt x \\{} - [f_j(x,t)c_j\phi^r_k(x)]_{x=y_k}^{x=y_{k+1}}. \end{multlined} \label{eqn:hash}
\end{align}
Now, \(f_j^W(x,t)\in W\) so, on \(\mathcal D_k\), it can be expressed as \(u_{k,j}(x,t):=\boldsymbol a_{k,j}(t) \boldsymbol \phi_k(x)\tr{}\), which we now substitute into (\ref{eqn:hash})
%\begin{align*}
%	\begin{multlined}[t]\int_{x\in\calD_k}\cfrac{\wrt}{\wrt t} \boldsymbol a_{k,j}(t) \boldsymbol \phi^k(x)\tr{}\phi^r_k(x)\wrt x = \int_{x\in\calD_k}\sum_{i\in\calS} \boldsymbol a_{k,i}(t) \boldsymbol \phi^k(x)\tr{}T_{ij}\phi^r_k(x)\wrt x  
%	%
%	\\ {}+  \int_{x\in\calD_k}\boldsymbol a_{k,j}(t) \boldsymbol \phi^k(x)\tr{}c_j\cfrac{\wrt}{\wrt x}\phi^r_k(x)\wrt x - c_j[f_j(x,t)\phi^r_k(x)]_{x=y_k}^{x=y_{k+1}}.\end{multlined}
%\end{align*}
and repeat this for all test functions \(\phi^r_k(x)\), \(r=1,...,N_k\), to get the following system of equations,
\begin{align}
	\begin{multlined}[t]\int_{x\in\calD_k}\cfrac{\wrt}{\wrt t} \boldsymbol a_{k,j}(t) \boldsymbol \phi_k(x)\tr{}\boldsymbol \phi_k(x)\wrt x = \int_{x\in\calD_k}\sum_{i\in\calS} \boldsymbol a_{k,i}(t) \boldsymbol \phi_k(x)\tr{}T_{ij}\boldsymbol\phi_k (x)\wrt x  
	%
	\\{}+  \int_{x\in\calD_k}\boldsymbol a_{k,j}(t) \boldsymbol \phi_k(x)\tr{}c_j\cfrac{\wrt}{\wrt x}\boldsymbol\phi_k(x)\wrt x - c_j[f_j(x,t) \boldsymbol\phi_k(x)]_{x=y_k}^{x=y_{k+1}},\quad k\in\mathbb Z. \end{multlined} \label{eqn:DG system}
\end{align}

\section{Mass, Stiffness, and Flux}
For \(k\in\mathbb Z\), define local \textit{mass} and \textit{stiffness} matrices \(\bs M_k\) and \(\bs G_k\) by 
\[\bs M_k := \int_{x\in\calD_k}\boldsymbol \phi_k(x)\tr{}\boldsymbol \phi_k(x)\wrt x,\quad \bs G_k := \int_{x\in\calD_k}\boldsymbol \phi_k(x)\tr{}\cfrac{\wrt}{\wrt x} \boldsymbol \phi_k(x)\wrt x.\]
%and global mass and stiffness matrices by
%\begin{align*}
%	M &:= \left[\begin{array}{ccc} M_1 & & \\ & \ddots & \\ & & M_K \end{array}\right] = \int_{x\inb}\boldsymbol \phi(x)\tr{}\boldsymbol \phi(x)\wrt x,
%	\\G &:= \left[\begin{array}{ccc} G_1 & & \\ & \ddots & \\ & & G_K \end{array}\right] = \int_{x\inb}\boldsymbol \phi(x)\tr{}\cfrac{\wrt}{\wrt x}\boldsymbol \phi(x)\wrt x,
%\end{align*}
We can write (\ref{eqn:DG system}) as 
\begin{align}
	&\cfrac{\wrt}{\wrt t} \boldsymbol a_{k,j}(t) \bs M_k = \sum_{i\in\calS} \boldsymbol a_{k,i}(t)\bs M_kT_{ij} 
	%
	+  c_j\boldsymbol a_{k,j}(t) \bs G_k - c_j[f_j(x,t)\boldsymbol\phi_k(x)]_{x=y_k}^{x=y_{k+1}}.\label{eqn:DG matrix}
\end{align}

It remains to approximate the \textit{flux}, \(f_j(x,t)\) at the cell edges \(y_k,\,k\in\mathbb Z\), so that we may evaluate the terms \([f_j(x,t)\phi^r_k(x)]_{x=y_k}^{x=y_{k+1}}\), \(r=1,...,N_k,\,k\in\mathbb Z\). This is the key for DG -- it joins the local approximations on each cell \(\calD_k\), into a global approximation on the whole domain of approximation. The flux is the instantaneous rate (with respect to time) at which density moves across the boundaries \(y_k,\,k\in\mathbb Z\). There are different choices for the flux, and we refer the reader to \citep{c99,nodalDGBook}, and references therein, for some discussion of the topic. Here, we choose the \textit{upwind} scheme, which, as we shall see, closely resembles the flux terms from the generator \(\mathbb B\). The approximate flux, also known as the \textit{numerical flux}, is given by 
\[f^*_j(x,t) = sign(c_j)\lim_{\varepsilon\to0^+}u_j(x-\varepsilon c_j,t),\]
at each \(x=y_k,\,k\in\mathbb Z\). 
Intuitively, the upwind flux takes the value of the density immediately on the upwind side of each \(y_k\).

>>>>

%!TEX root = ../thesis.tex
\chapter{Approximating fluid queues with the discontinuous Galerkin method\label{ch:galerkin}}

In this chapter we introduce the discontinuous Galerkin (DG) method applied to fluid queues to approximate the operator-analytic expressions for fluid-fluid queues in \cite{bo2014}. 
\begin{center}
	\begin{minipage}{0.8\textwidth}
		\textit{Apart from Sections~\ref{sec: other applications}-\ref{sec: slope limiting}, this chapter has been taken from Section~4 and 5 of \cite{blnos2022} with only minor changes, such as notations, so that this chapter is consistent with the rest of the thesis. I am a co-author of the paper \cite{blnos2022}. %The conceptualisation of \cite{blnos2022} was originally by Vikram Sunkara, Nigel Bean and Giang Nguyen, and the original coding was done by Vikram Sunkara. I made significant contributions to Section~3 of the paper, expressing the operator-theoretic expressions to use the same partition as the approximation scheme. I contributed Sections~4.4 and 5.1. I extended the numerical experiments in Section~6 to higher orders and made all the plots in Section~6. Appendix~A is also my original work. I did a significant proportion of the writing of the manuscript and addressed the reviewers comments and also developed code for the numerical experiments.
		}
	\end{minipage}
	\end{center}
% The DG method for B 
    % direct copy paste from the paper
% Problems with DG, non-neg/oscillations
    % solutions: filtering and limiting
% Approximating R: projection, interpolation, cell averages (eluding to QBD-RAP and unif method)
% Approximating Psi (and the stationary distribution, but this isnt really necessary)
    % constructing D
    % solving the matrix-Riccati equation, iteration/algorithm.

@@

% \section{Discontinuous Galerkin Approximation of the Generator of a Fluid Queue}
	% \label{sec:DG}
Discontinuous Galerkin (DG) methods can be used to approximate the solutions to systems of partial differential equations (PDEs). For a most readable and practical text on DG methods we refer the reader to \cite{nodalDGBook}. Here we describe how one can apply the DG method to approximate fluid queues and fluid-fluid queues. In finite-element methods, such as the DG method, the domain of approximation is partitioned into intervals, referred to individually as \textit{cells} and collectively as a \textit{mesh}. On each cell we construct a finite-dimensional smooth Sobolev space using piecewise-polynomial basis functions, and then project the partial differential equations onto this space. This projection leads to a new system of equations which represent a \textit{weak form} of the original system of PDEs. Next, we approximate the \textit{flux operator} which moves probability from one cell to another, in a manner similar to the underlying principle of a finite-volume approximation. This creates a system of ordinary differential equations which is known as a \emph{semi-discrete system} (discrete in the spatial variable but continuous in time). We can solve the semi-discrete system to approximate transient distributions of the fluid queue or, as we do in Section~\ref{sec:DGSFFM}, we can extract and manipulate an approximation to the generator of the fluid queue to approximate performance measures of fluid-fluid queues. The DG method conserves probability. Here we construct the DG approximation to the matrix of operators \(\mathbb B\) which we use later to construct a DG approximation to \(\mathbb D(s)\) then \(\mathbb\Psi(s)\), and ultimately the stationary distribution of a stochastic fluid-fluid queue.

\section{The partial differential equation}
We start by introducing the PDE from which we will extract the approximation to the generator \(\mathbb B\).

Let $f_i(x,t)$ be the joint density of $(X_t, \varphi_t)$: 
	% 
	\begin{align*} 
	    f_i(x,t) := \cfrac{\partial}{\partial x} 	\mathbb{P}\left(X(t) \leq x, \varphi(t) = i\right),\quad 0<x<b,\, i \in\calS,
	\end{align*} 
which satisfies the system of partial differential equations 
%
\begin{align*}
\cfrac{\partial}{\partial t} f_i(x,t)& = \sum_{j\in \mathcal{S}}  f_j(x,t)T_{ji} - c_i \cfrac{\partial}{\partial x} f_i(x,t), \quad 0<x<b,\, i\in\mathcal S,
\end{align*}
% 
subject to suitable boundary conditions. In matrix form, 
\begin{align}
\cfrac{\partial}{\partial t} \boldsymbol f(x,t) &= \boldsymbol f(x,t)\bs T -  \cfrac{\partial}{\partial x}\boldsymbol f(x,t)\widehat{\bs C}\label{eqn:pde_density_matrix}, 
\end{align}
where \(\boldsymbol f(x,t) = \left(f_i(x,t)\right)_{i\in\mathcal S}\) is a row-vector. 
% This system of PDEs is closely related to the generator \(\mathbb B\); \(\boldsymbol \mu(t)\) satisfies the operator differential equation
% \begin{align*}
% 	\cfrac{\wrt}{\wrt t}\boldsymbol \mu(t)(\wrt x)  = \boldsymbol \mu(t)\mathbb B(\wrt x) = \boldsymbol f(x,t)\bs T\wrt x - \cfrac{\partial}{\partial x}\boldsymbol f(x,t)\bs C\wrt x,
% \end{align*}
% on the interior of the space \([0,\infty)\). Thus, by approximating the operator on the right-hand side of Equation~(\ref{eqn:pde_density_matrix}) we can approximate the infinitesimal operator \(\mathbb B\). The DG method does exactly this, by approximating the operator with a matrix.

\section{Cells, test functions, and weak formulation}
To begin with, consider an unbounded first fluid level \(\{\ddot X_t,t\geq0\}\), \(\ddot X_t\in(-\infty,\infty)\). We will eventually truncate this space so that we have a finite dimensional approximation; however, this requires a discussion on boundary conditions which we save for later. Let \(\mathcal D_{k,i} = [y_k,y_{k+1}),\, k\in\mathbb Z\) for \(i\in\calS_+\cup\calS_0\) and \(\mathcal D_{k,i} = (y_k,y_{k+1}],\, k\in\mathbb Z\) for \(i\in\calS_-\), \(y_k<y_{k+1}\), partition the domain \((-\infty,\infty)\). We call the \(\calD_{k,i}\) \textit{cells} and define \(\Delta_k=y_{k+1}-y_k\).

For each \(k\) we choose \(N_k\) linearly independent functions \(\{\phi^r_k\}_{r=1}^{N_k}\), compactly supported on \(\calD_{k,i}\) (i.e.~\(\phi^r_k(x)=0\) for \(x\notin\calD_{k,i}\)) to form a basis for the space \(W_k\), in which we formulate the approximation. Here, as is standard in DG methods \citep{nodalDGBook}, we take \(\{\phi^r_k\}_{r=1}^{N_k}\) to be the space of polynomials of degree \(N_k-1\). It is convenient in this work to take \(\{\phi^r_k\}_{r=1}^{N_k}\) as a basis of Lagrange interpolating polynomials defined by the Gauss-Lobatto quadrature points, since our approximations inherit nice properties from this \citep{nodalDGBook}. However, the constructions presented here are general, and any basis can be used. For the sake of illustration, the reader may think of \(\{\phi^r_k\}_{r=1}^{N_k}\) as the Lagrange polynomials. On each cell \(\mathcal D_{k,i}\) we approximate 
\[f_i(x,t)\approx u_{k,i}(x,t)=\sum\limits_{r=1}^{N_k}a_{k,i}^r(t)\phi^r_k(x),\] 
where \(a_{k,i}^r(t)\) are yet-to-be-determined time-dependent coefficients. We refer to \(u_{k,i}\) as the \textit{local} approximation on cell \(k\), while the \textit{global} approximation is given by \(\sum\limits_{k\in\mathbb Z}u_{k,i}\) on the whole domain. The whole approximation space is \(\bigoplus\limits_{k\in\mathbb Z} W_k\).

@@

The DG method proceeds by first considering the \textit{weak-formulation} of the PDE, which is constructed from the \textit{strong-form} of the PDE, Equation~(\ref{eqn:pde_density_matrix}). In general, to construct the weak-form we need a set of test functions, say \(W\). Now, take the strong form of the PDE, multiply it by some test function \(\psi(x)\in W\), integrate with respect to \(x\), and apply integration by parts to the derivative with respect to \(x\), to get 
\begin{align}
\begin{multlined}[t]
	\int_{x\in\mathbb R}\cfrac{\partial}{\partial t} f_j(x,t)\psi(x)\wrt x = \int_{x\in\mathbb R} \sum_{i\in\calS}f_i(x,t)T_{ij}\psi(x)\wrt x 
	%
	+  \int_{x\in\mathbb R} f_j(x,t)c_j\cfrac{\wrt}{\wrt x}\psi(x)\wrt x \\{}- [ f_j(x,t)c_j\psi(x)]_{x=-\infty}^{x=\infty}, \end{multlined}\label{eqn:weak form}
\end{align}
for \(j\in\calS\). It is common to choose \(W\) such that \(\psi(-\infty)=\psi(\infty)=0\), in which case the last term on the right is zero. Requiring (\ref{eqn:weak form}) to hold for every \(\psi\in W\) gives the weak-formulation of the PDE. For a sufficiently rich set of test functions \(W\) the weak and strong forms of the PDE are equivalent. Solutions to (\ref{eqn:weak form}) are known as \textit{weak} solutions and generalise the concept of a solution of the PDE. For example, this may allow discontinuities with respect to \(x\) in the solution -- something which is ill-defined for the strong form.

For the purpose of DG, we take the set of test functions to be \(W = \bigoplus\limits_{k\in\mathbb Z} W^k\), the same as the set of basis functions of our solution space. Proceeding as described above, the weak formulation is 
\begin{align*}
	\begin{multlined}[t]\int_{x=y_k}^{y_{k+1}}\cfrac{\partial}{\partial t} f_j(x,t)\phi^r_k(x)\wrt x = \int_{x=y_k}^{y_{k+1}}\sum_{i\in\calS}  f_i(x,t)T_{ij}\phi^r_k(x)\wrt x  
	%
	+  \int_{x=y_k}^{y_{k+1}} f_j(x,t)c_j\cfrac{\wrt}{\wrt x}\phi^r_k(x)\wrt x \\ {}- [f_j(x,t)c_j\phi^r_k(x)]_{x=y_k}^{x=y_{k+1}}, \end{multlined}
\end{align*}
since \(\phi^r_k\) is compactly supported on \(\calD_{k,j}\), for all \(j\in\mathcal S,\,r\in\mathcal N_k\), \(k\in\mathbb Z.\) Now, note that any function \(g(x)\) can be decomposed as \(g(x) = g^{W}(x)+g^\perp(x)\) where \(g^{W}\in W\) and \(g^\perp \in W^\perp\), and \(W^\perp\) is the orthogonal complement of \(W\). Since \(g^\perp\) is orthogonal to \(W\), \(\displaystyle\int_{x}g^\perp(x)\phi^r_k(x)\wrt x=0\) for \(r\in\mathcal N_k,\,k\in\mathbb Z\). Also, note that \(\cfrac{\wrt}{\wrt x}\phi^r_k(x)\in W\). Using this, we can write 
\begin{align*}
	\begin{multlined}[t]\int_{x=y_{k}}^{y_{k+1}}\cfrac{\partial}{\partial t} \left(f_j^W(x,t)+f_j^\perp(x,t)\right)\phi^r_k(x)\wrt x 
	= \int_{x=y_k}^{y_{k+1}}\sum_{i\in\calS} \left(f_i^W(x,t)+f_i^\perp(x,t)\right)T_{ij}\phi^r_k(x)\wrt x  
	%
	\\ {}
	+  \int_{x=y_k}^{y_{k+1}} \left(f_j^W(x,t)+f_j^\perp(x,t)\right)c_j\cfrac{\wrt}{\wrt x}\phi^r_k(x)\wrt x - [f_j(x,t)c_j\phi^r_k(x)]_{x=y_k}^{x=y_{k+1}}, \end{multlined}
\end{align*}
	which is equivalent to
\begin{align}
	\begin{multlined}[t]\int_{x=y_k}^{y_{k+1}}\cfrac{\partial}{\partial t} f_j^W(x,t)\phi^r_k(x)\wrt x = \int_{x=y_k}^{y_{k+1}}\sum_{i\in\calS} f_i^W(x,t)T_{ij}\phi^r_k(x)\wrt x  
	%
	\\{} + \int_{x=y_k}^{y_{k+1}} f_j^W(x,t)c_j\cfrac{\wrt}{\wrt x}\phi^r_k(x)\wrt x - [f_j(x,t)c_j\phi^r_k(x)]_{x=y_k}^{x=y_{k+1}}. \end{multlined} \label{eqn:hash}
\end{align}
Now, \(f_j^W(x,t)\in W\) so, on \(\mathcal D_{k,j}\), it can be expressed as \(u_{k,j}(x,t):=\boldsymbol a_{k,j}(t) \boldsymbol \phi_k(x)\tr{}\), which we now substitute into (\ref{eqn:hash})
%\begin{align*}
%	\begin{multlined}[t]\int_{x\in\calD_k}\cfrac{\wrt}{\wrt t} \boldsymbol a_{k,j}(t) \boldsymbol \phi^k(x)\tr{}\phi^r_k(x)\wrt x = \int_{x\in\calD_k}\sum_{i\in\calS} \boldsymbol a_{k,i}(t) \boldsymbol \phi^k(x)\tr{}T_{ij}\phi^r_k(x)\wrt x  
%	%
%	\\ {}+  \int_{x\in\calD_k}\boldsymbol a_{k,j}(t) \boldsymbol \phi^k(x)\tr{}c_j\cfrac{\wrt}{\wrt x}\phi^r_k(x)\wrt x - c_j[f_j(x,t)\phi^r_k(x)]_{x=y_k}^{x=y_{k+1}}.\end{multlined}
%\end{align*}
and repeat this for all test functions \(\phi^r_k(x)\), \(r=1,...,N_k\), to get the following system of equations,
\begin{align}
	\begin{multlined}[t]\int_{x=y_k}^{y_{k+1}}\cfrac{\wrt}{\wrt t} \boldsymbol a_{k,j}(t) \boldsymbol \phi_k(x)\tr{}\boldsymbol \phi_k(x)\wrt x = \int_{x=y_k}^{y_{k+1}}\sum_{i\in\calS} \boldsymbol a_{k,i}(t) \boldsymbol \phi_k(x)\tr{}T_{ij}\boldsymbol\phi_k (x)\wrt x  
	%
	\\{}+  \int_{x=y_k}^{y_{k+1}}\boldsymbol a_{k,j}(t) \boldsymbol \phi_k(x)\tr{}c_j\cfrac{\wrt}{\wrt x}\boldsymbol\phi_k(x)\wrt x - c_j[f_j(x,t) \boldsymbol\phi_k(x)]_{x=y_k}^{x=y_{k+1}},\quad k\in\mathbb Z. \end{multlined} \label{eqn:DG system}
\end{align}

\section{Mass, stiffness, and flux matrices}
For \(k\in\mathbb Z\), define local \textit{mass} and \textit{stiffness} matrices \(\bs M_k\) and \(\bs G_k\) by 
\[\bs M_k := \int_{x\in\calD_k}\boldsymbol \phi_k(x)\tr{}\boldsymbol \phi_k(x)\wrt x,\quad \bs G_k := \int_{x\in\calD_k}\boldsymbol \phi_k(x)\tr{}\cfrac{\wrt}{\wrt x} \boldsymbol \phi_k(x)\wrt x.\]
%and global mass and stiffness matrices by
%\begin{align*}
%	M &:= \left[\begin{array}{ccc} M_1 & & \\ & \ddots & \\ & & M_K \end{array}\right] = \int_{x\inb}\boldsymbol \phi(x)\tr{}\boldsymbol \phi(x)\wrt x,
%	\\G &:= \left[\begin{array}{ccc} G_1 & & \\ & \ddots & \\ & & G_K \end{array}\right] = \int_{x\inb}\boldsymbol \phi(x)\tr{}\cfrac{\wrt}{\wrt x}\boldsymbol \phi(x)\wrt x,
%\end{align*}
We can write (\ref{eqn:DG system}) as 
\begin{align}
	&\cfrac{\wrt}{\wrt t} \boldsymbol a_{k,j}(t) \bs M_k = \sum_{i\in\calS} \boldsymbol a_{k,i}(t)\bs M_kT_{ij} 
	%
	+  c_j\boldsymbol a_{k,j}(t) \bs G_k - c_j[f_j(x,t)\boldsymbol\phi_k(x)]_{x=y_k}^{x=y_{k+1}}.\label{eqn:DG matrix}
\end{align}

It remains to approximate the \textit{flux}, \(f_j(x,t)\) at the cell edges \(y_k,\,k\in\mathbb Z\), so that we may evaluate the terms \([f_j(x,t)\phi^r_k(x)]_{x=y_k}^{x=y_{k+1}}\), \(r=1,...,N_k,\,k\in\mathbb Z\). This is the key for DG -- it joins the local approximations on each cell into a global approximation on the whole domain of approximation. The flux is the instantaneous rate (with respect to time) at which density moves across the boundaries \(y_k,\,k\in\mathbb Z\). There are different choices for the flux, and we refer the reader to \citep{c99,nodalDGBook}, and references therein, for some discussion of the topic. Here, we choose the \textit{upwind} scheme, which, as we shall see, closely resembles the flux terms from the generator \(\mathbb B\). The approximate flux, also known as the \textit{numerical flux}, is given by 
\[f^*_j(x,t) = sign(c_j)\lim_{\varepsilon\to0^+}u_j(x-\varepsilon c_j,t),\]
at each \(x=y_k,\,k\in\mathbb Z\). 
Intuitively, the upwind flux takes the value of the density immediately on the upwind side of each \(y_k\).

Denote by \(x^-\) and \(x^+\) the left and right limits at \(x\), respectively. Assume first \(c_j>0\), then 
\begin{align*}
	-c_j[f_j(x,t)\phi^r_k(x)]_{x=y_k}^{x=y_{k+1}} & \approx -c_j[f_j^*(x,t)\phi^r_k(x)]_{x=y_k}^{x=y_{k+1}}
	\\& = -c_jf_j^*(y_{k+1},t)\phi^r_k(y_{k+1}) + c_jf_j^*(y_{k},t)\phi^r_k(y_{k})
	\\& = -c_ju_j(y_{k+1}^-,t)\phi^r_k(y_{k+1}) + c_ju_j(y_{k}^-,t)\phi^r_k(y_{k})
	\\& = -c_ju_{k,j}(y_{k+1}^-,t)\phi^r_k(y_{k+1}) + c_ju_{k-1,j}(y_{k}^-,t)\phi^r_k(y_{k})
	\\& = -c_j\boldsymbol a_{k,j}(t)\boldsymbol \phi_k(y_{k+1}^-)\tr{}\phi^r_k(y_{k+1}) + c_j\boldsymbol a_{k-1,j}(t)\boldsymbol \phi_{k-1}(y_{k}^-)\tr{}\phi^r_k(y_{k}).
\end{align*}
In matrix form,  
\begin{align*}
	-c_j[f_j(x,t)\boldsymbol\phi_k(x)]_{x=y_k}^{x=y_{k+1}} & \approx -c_j[f_j^*(x,t)\boldsymbol\phi_k(x)]_{x=y_k}^{x=y_{k+1}}
	\\& = -c_j\boldsymbol a_{k,j}(t)\boldsymbol \phi_k(y_{k+1}^-)\tr{}\boldsymbol\phi_k(y_{k+1}) + c_j\boldsymbol a_{k-1,j}(t)\boldsymbol \phi_{k-1}(y_{k}^-)\tr{}\boldsymbol\phi_k(y_{k})
	\\& = c_j\boldsymbol a_{k,j}(t)F_j^{k,k}+c_j\boldsymbol a_{k-1,j}(t)F_j^{k-1,k},
\end{align*}
where, for \(j\in \calS\) with \(c_j>0\), we define \(\bs F_+^{k,k}=\bs F_j^{k,k} := -\boldsymbol \phi_k(y_{k+1}^-)\tr{}\boldsymbol\phi_k(y_{k+1}),\,k\in\mathbb Z\) and \(\bs F_j^{k-1,k} := \boldsymbol \phi_{k-1}(y_{k}^-)\tr{}\boldsymbol\phi_k(y_{k}),\,k\in\mathbb Z\).
